---
title: marketr package released to CRAN
author: Chris Umphlett
date: '2020-02-15'
slug: marketr-package-released-to-cran
categories: []
tags: []
description: ''
externalLink: ''
---

For the last few months I have been working on turning some previous work into an R package `marketr`. I needed to learn a quite a bit to develop it as a working package posted to GitHub; and in turn, do even more to get it on to CRAN. Once you understand the end product there are a lot of things I could have done differently in the development to help next time.

`devtools` package gives a lot of great feedback to prepare you for CRAN submission. I ended up submitting to CRAN and only having to make some fairly minor changes to then get acceptance of the package in my second submission.

I developed the code behind `marketr` in order to save myself time when doing ad hoc calculations of marketing metrics along different dimensions. My team at that time had some standard hierarchical reporting but we were supplementing that with analysis of other slices of the data--- going deeper, or grouping by multiple dimensions. `marketr` allows me to quickly do these calculations with whatever combination of dimensions I specify.

Putting it into a package helped ensure consistency of calculation, especially for others who were inexperienced R users. They no longer needed to try and adapt my code, or have the possibility of the calculation being incorrect. `marketr` is useful regardless of your speed or experience with R if you are working with these metrics.

Below I've adapted some of the material from the README and vignette to explain what the package does and show how it does it. Visit the [GitHub](https://github.com/chrisumphlett/marketr) to learn more or contribute.


## Introduction to marketr

`marketr` facilitates tidy calculation of popular quantitative marketing metrics (like Customer Experience Index and Net Promoter Score). By "tidy", I am referring to the usage of the tidyverse packages and methodology for organizing and analyzing data. The package is designed so that beginning R users can calculate these metrics, along many dimensions, without needing to learn much R syntax. It is also helpful for more experienced programmers to do these calculations quickly.

## Installation
`install.packages("marketr")` provides the current verision from CRAN.

The development version can be installed from GitHub: `devtools::install_github("chrisumphlett/marketr")`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6)

```

## Generate survey response data

To demonstrate the basic usage I will create simulated survey response data. `needs`, `ease` and `emotion` are the columns that make up CXi; `nps_question` is used for NPS; `grps` and `months` will show how these metrics can be calculated along categorical features and/or trended over time.

```{r gen_data}
library(marketr)
library(dplyr)
library(magrittr)
library(ggplot2)

needs <- sample(2:5, 1000, replace = T)
ease <- sample(2:5, 1000, replace = T)
emotion <- sample(2:5, 1000, replace = T)
nps_question <- sample(3:10, 1000, replace = T)
grps <- c("a", "b", "c")
months <- sample(1:12, 1000, replace = T)

survey_data <- tibble::as_tibble(cbind(needs, ease, emotion, nps_question, grps, months)) %>%
  mutate(month = as.numeric(months))

head(survey_data)
```

## Calculating CXi

*Customer Experience Index* (CXI) was [developed by Forrester](https://go.forrester.com/analytics/cx-index/). Per Forrester, CXi "measures how successfully a company delivers customer experiences that create and sustain loyalty." 

It involves scoring three questions, each with a likert scale response, and then averaging those scores together. Below, four calculations are done using two different functions.

```{r cxi}
# Overall CXi
cxi_calc(survey_data) %>% knitr::kable()

## CXi by group
cxi_calc(survey_data, grps, cx_high = 4, cx_low = 2) %>% knitr::kable()

# Overall CXi trend
cxi_trend(survey_data, month) %>% knitr::kable() 

# Overall CXi trend by group - plotted
cxi_trend(survey_data, month, grps, cx_high = 4, cx_low = 2, min_surveys = 1, avg_surveys = 0) %>% 
  ggplot(aes(x = month, y = cxi)) +
  geom_line() +
  facet_wrap(grps ~ ., nrow = 3)
```


## Calculating NPS

*Net Promoter Score* (NPS) was originally developed by Fred Reichheld and now is owned by [Bain Company and Satmetrix Systems](https://www.netpromoter.com/). The [Wikipedia page](https://en.wikipedia.org/wiki/Net_Promoter) is another good source of information. According to Wikipedia it "is a management tool that can be used to gauge the loyalty of a firm's customer relationships."

The calculation requires a single question with a ten-point scale. Like CXi it is not difficult to do manually; the package enables deeper analysis.Below, four calculations are done using two different functions.

```{r NPS}
# Overall NPS
nps_calc(survey_data) %>% knitr::kable()

## NPS by group
nps_calc(survey_data, grps) %>% knitr::kable()

# Overall NPS trend
nps_trend(survey_data, month) %>% knitr::kable()

# Overall NPS trend by group - plotted
nps_trend(survey_data, month, grps, min_surveys = 1, avg_surveys = 0) %>% 
  ggplot(aes(x = month, y = nps)) +
  geom_line() +
  facet_wrap(grps ~ ., nrow = 3)
```
